A Deep Learning Approach for Gait Event Detection from a Single Shank-Worn IMU: Validation in Healthy and Neurological Cohorts

### 2022-04-21 paper/MDPI_template/article.tex
Cleaned working directory. Setup __obsolete__ directory.

% Authors, for the paper (add full first names)
\Author{Robbin Romijnders$^{1}$*\orcidA{}, Elke Warmerdam$^{2}$, Clint Hansen$^{1}$*, Gerhard Schmidt$^{3}$ and Walter Maetzler$^{1}$}

@@

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{%
$^{1}$ \quad Department of Neurology, Kiel University, 24105 Kiel, Germany; r.romijnders@neurologie.uni-kiel.de (R.R.), c.hansen@neurologie.uni-kiel.de (C.H.), w.maetzler@neurologie.uni-kiel.de (W.M.)\\
$^{2}$ \quad xxxxx, xxxxx, xxxxx Homburg, Germany; elke.warmerdam@uni-saarland.de (E.W.)\\
$^{3}$ \quad Institute of Electrical Engineering and Information Technology, Faculty of Engineering, Kiel University, 24143 Kiel, Germany; gus@tf.uni-kiel.de (G.S.)}

@@

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{Many algorithms use 3D accelerometer and/or gyroscope data from IMU sensors to detect gait events (i.e., initial and final contact). However, these algorithms often require knowledge about sensor orientation and use empirically derived thresholds. As alignment cannot always be controlled for in ambulatory assessments, methods are needed that require little knowledge on sensor location and orientation, e.g., a convolutional neural network-based deep learning model. Therefore, 160 participants from healthy and neurologically diseased cohorts walked 5 meter distances at slow, preferred, and fast walking speed, while data were collected from IMUs on the left and right ankle and shank. Gait events were detected and stride parameters were extracted using a deep learning model and an optoelectronic motion capture (OMC) system for reference. Results showed a high detection rate for both initial and final contact detection across the IMUs (recall~$\ge 92\%$, precision~$\ge 97\%$). Time agreement was excellent, as witnessed median time error (0.005 s) and corresponding inter-quartile range (0.020~s). The extracted stride-specific parameters were in good agreement with parameters derived from the OMC system (maximum mean difference 0.003~s and corresponding maximum limits of agreement (-0.049, 0.051)~s for $95\%$ confidence level). Thus, the deep learning approach was considered a valid approach for detecting gait events and extracting stride-specific parameters with little knowledge on exact IMU location and orientation in conditions with and without walking pathologies due to neurological diseases.}

% Keywords
\keyword{gait; inertial measurement units; deep learning}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction \label{sec:Introduction}}
Gait deficits are common in older adults and possibly reflect the presence of an underlying neurodegenerative disease \cite{Snijders2007,Hodgins2008}. For example, conversion to Parkinson's Disease \cite{DelDin2019} or from mild cognitive impairment to Alzheimer's Disease \cite{Koenig2017,Bertoli2018} are linked with changes in spatiotemporal gait parameters. Similarly, temporal gait parameters are different for stroke patients \cite{SchroederVon1995,Mohan2021} and patients with multiple sclerosis \cite{Griskevicius2016,Flachenecker2019} when compared to healthy controls. To objectively quantify gait deficits, stride-specific parameters such as stride time or stride length are often used \cite{Hannink2016}. The beginning and end of a stride are determined from two successive initial contacts (ICs) of the same foot \cite{Perry2010,Whittle2012}. The IC is when the foot contacts the ground, and together with the instant at which the foot leaves the ground (final contact, FC), each stride can be divided in a stance and swing phase \cite{Rueterbories2010,Bruening2014}.  The events of IC and FC, also referred to as \emph{gait events}, are commonly determined using force or pressure measuring devices \cite{Bruening2014}, or marker-based optoelectronic motion capture systems (henceforth referred to as the \emph{marker-based} system or method) \cite{Chiari2005,Topley2020}. These systems are relatively expensive, and restricted to usage in expertise laboratories \cite{Iosa2016,Jarchi2018}. As there is increasing evidence that gait measured in the lab does not reflect daily-life gait \cite{Hillel2019,Warmerdam2020,Atrsaei2021}, there is increasingly more interest in measurement systems that allow for continuous gait analysis in ambulatory settings. Therefore, the use of inertial measurement units (IMUs) is especially attractive, as these can be used to measure gait in ecologically valid environments, such as the home environment, thereby painting a more complete picture of health status \cite{DelDin2016,Shah2020} and providing clinical information that is complementary to standardized lab-based assessments \cite{Warmerdam2020,Fasano2020,Atrsaei2021,Corra2021}.

Previous research suggests that gait event detection is more accurate using an IMU worn on a lower limb (e.g., shank or foot) compared to an IMU worn on the low back \cite{BenMansour2015,Storm2016,Panebianco2018}. Now, in order to get from abstract IMU sensor readings to clinically relevant gait parameters (e.g., from accelerations and angular velocities to stride times) \cite{Hannink2016}, different algorithmic approaches have been developed in the last twenty years of clinical gait research. A recent study has evaluated a cross-section of these algorithms for different sensor locations on the lower leg and foot \cite{Niswander2021}. The algorithms were categorized according to which signals were analyzed, for example the angular velocity about the medio-lateral axis, or the accelerations along vertical and antero-posterior axis. This means the sensor readings need to be linked with the anatomical axes, that is, one needs to know which sensor axis aligns with for example the medio-lateral axis. In most approaches, it is simply assumed that due to sensor attachment the sensor axis aligns roughly with the anatomical axis of interest (\cite{Salarian2004,Sabatini2005,Jasiewicz2006,Catalfamo2010,Trojaniello2014,Maqbool2016,Romijnders2021}) or an additional calibration procedure (e.g., \cite{Ferraris1995}) is required (\cite{Greene2010,Niswander2021}). In ambulatory assessments however, study participants often attach the sensor themselves, and therefore the sensor location and alignment cannot be controlled for. Furthermore, it is unlikely that each time the sensor is (re-)attached study participants, especially those with gait deficits, perform a calibration procedure that usually consists of holding a pre-defined pose and performing some known movement sequences (TODO: reference?! Kong? Seel?).

Taken together, this drives the need for an approach that is invariant to sensor orientation, and is applicable across a variety of pathological gait patterns. As pointed out by \cite{Hannink2016}, in the field of image analysis similar requirements have been successfully addressed by algorithms that share a common underlying methodology referred to as deep learning \cite{LeCun2015,TerHaarRomeny2019}. Recent applications of deep learning algorithms have already shown improved performance in detecting gait events from marker-based motion capture when compared to conventional, often heuristics-based, algorithms \cite{Kidzinski2019,Lempereur2020,Filtjens2020}. Another study used a deep learning approach to detect gait events from either three IMUs (worn on the low back, and both ankles) or a single IMU (worn on the low back), and showed that the time error was considerably smaller for the deep learning algorithm than for a commonly applied wavelet-based approach \cite{Gadaleta2019}.

The aim of this study is to extend these works by validating this deep learning approach for detecting gait events in a heterogeneous cohort of healthy and neurologically diseased adults considering a single IMU setup, worn on the lower leg.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Materials and Methods \label{sec:Materials_and_Methods}}
\subsection{Data Collection \label{subsec:Data_Collection}}
Gait analyses were performed in the Universit\"{a}tsklinikum Schleswig-Holstein (UKSH) campus Kiel, Germany. The study \cite{Warmerdam2021} was approved by the ethical committee of the medical faculty at the UKSH (no: D438/18). In total, data from 160 participants were included for the current analysis, including data from young adults (YA; age: 18 - 60 years), older adults (OA; age: $>$60 years), people with Parkinson's Disease (PD; according to the UK Brain Bank criteria \cite{Gibb1988}), people with a recent ($<$4 weeks) symptomatic stroke (stroke), people with multiple sclerosis (MS; according to the McDonalds criteria \cite{Thompson2018}), people with chronic low back pain (cLBP), and people with other diagnoses that affected mobility (Table \ref{tab:demographics_data}). Inclusion criteria were an age of 18 years or older, and the ability to walk independently without a walking aid. Participants were excluded from the study with a Montreal Cognitive Assessment \cite{Nasreddine2005} score $<$15 and other movement disorders that affected mobility, as noticed by the clinical assessor.

@@

\begin{figure}[H]
	\begin{adjustwidth}{-\extralength}{0cm}
		\centering
		\includegraphics[width=13.5cm]{fig/methods_overall_workflow.pdf}
	\end{adjustwidth}
	\caption{Schematic depiction$^{1}$ of the current study. Study participants wore IMUs on the ankle and shanks, and reflective markers were adhered on the heel and toe of usual footwear (shown on the left). Marker data were used to obtain reference values for the timings of initial and final contacts (top), where accelerometer and gyroscope data from each tracked point were inputted to a neural network that predicted timings of the same initial and final contacts (bottom). \newline \noindent{\footnotesize{\textsuperscript{1} Picture from: \url{https://www.vecteezy.com/free-vector/man-walking }.}} \label{fig:methods_marker_vs_imu}}
\end{figure}

@@

For the current analysis data from four IMUs (Noraxon USA Inc., myoMOTION, Scottsdale, AZ, USA) were considered, namely those that were attached laterally above the left and right ankle joint and those attached proximally at the left and right shank. IMUs were secured to participants using elastic bands with a special hold for the IMU. Furthermore, reflective markers were attached on top of the usual foot wear at the heel and toe of both feet (Figure~\ref{fig:methods_marker_vs_imu}). Marker data were recorded using a twelve-camera optoelectronic motion capture system (Qualisys AB, G\"{o}teborg, Sweden) at a sampling frequency of 200 Hz. IMU data were recorded at the same sampling frequency, and both systems were synchronized using a TTL signal \cite{Warmerdam2021}. For some recordings, the sampling frequency of the IMUs was mistakenly set at 100 Hz, and therefore data from these trials were upsampled before further analysis.

\subsection{Data Pre-processing \label{subsec:Data_Pre-processing}}
\subsubsection{Marker Data \label{subsubsec:Marker_Data}}
First, data from both marker and IMU systems were cropped so that only data from within the 5 meter distance was considered. Any gaps in the marker data were then filled by interpolation making use of inter-correlations between markers \cite{Federolf2013,Gloersen2016}. The data were then low-pass filtered using a double pass 6th order Butterworth filter with a cut-off frequency of 20 Hz \cite{Kormylo1974,Racz2021}. The filtered data were differentiated to get velocity signals, and timings of ICs and FCs were determined from local maxima and minima in the heel and toe vertical velocity signals \cite{Pijnappels2001,OConnor2007}. Like in \cite{Carcreff2018,Romijnders2021}, all identified ICs and FCs were manually checked using Qualisys Track Manager 2018.1 software (Qualisys AB, Göteborg, Sweden), and corrected if necessary. The resulting annotated ICs and FCs were considered the \emph{true} events (also \emph{labels} or \emph{targets}), and were used as reference timings to derive stride-specific gait parameters.

\subsubsection{IMU Data \label{subsubsec:IMU_Data}}
The idea behind the deep learning approach was that a \emph{model} was trained to predict the likelihood of an IC and FC, given accelerometer and gyroscope data from a single IMU. The data from a single sensor channel, e.g., the acceleration in forward direction, or the angular velocity about the vertical direction, was denoted by $\mathbf{x}_{d} = \begin{bmatrix}
x_{d}[1] & x_{d}[2] & \cdots & x_{d}[N]
\end{bmatrix}^{\mathrm{T}}$, with $d$ referring to the $d$th sensor channel (i.e., $d = 1, \cdots, D$) and $n$ referring to the $n$th sample or time step (i.e., $n = 1, \cdots, N$). Data from all $D$ sensor channels together was then denoted by:
\begin{equation}
	\mathbf{X} = \begin{bmatrix}
		\rule[0ex]{0.5pt}{2.5ex} & \rule[0ex]{0.5pt}{2.5ex} &  & \rule[0ex]{0.5pt}{2.5ex}\\
		\mathbf{x}_{1} & \mathbf{x}_{2} & \cdots & \mathbf{x}_{D}\\
		\rule[0ex]{0.5pt}{2.5ex} & \rule[0ex]{0.5pt}{2.5ex} &  & \rule[0ex]{0.5pt}{2.5ex}
		\end{bmatrix} = \begin{bmatrix}
		x_{1}[1] & x_{2}[1] & & x_{D}[1]\\
		x_{1}[2] & x_{2}[2] & & x_{D}[2]\\
		\vdots & \vdots & \cdots & \vdots\\
		x_{1}[N] & x_{2}[1] & & x_{D}[N]\end{bmatrix}, \quad \mathbf{X} \in \mathbb{R}^{N \times D}
	\label{eqn:train_data_single_instance}
\end{equation}

Likewise, the \emph{labels} were denoted by:
\begin{equation}
	\mathbf{y}_{\mathrm{IC}} = \begin{bmatrix}
		y_{\mathrm{IC}}[1]\\
		y_{\mathrm{IC}}[2]\\
		\vdots\\
		y_{\mathrm{IC}}[N]
		\end{bmatrix}, \quad \mathbf{y}_{\mathrm{FC}} = \begin{bmatrix}
		y_{\mathrm{FC}}[1]\\
		y_{\mathrm{FC}}[2]\\
		\vdots\\
		y_{\mathrm{FC}}[N]
		\end{bmatrix}, \quad y_{\mathrm{IC}/\mathrm{FC}}[n] \in [0, 1]
	\label{eqn:train_labels_single_instance}
\end{equation}

The model was then iteratively trained to learn a mapping $h_{\mathbf{\Theta}}(\mathbf{X}): \mathbf{X} \rightarrow \mathbf{y}$, where $h_{\mathbf{\Theta}}$ was also referred to as the \emph{hypothesis} which was parameterized by the weights and biases, collectively denoted by $\mathbf{\Theta}$.

All participant data were split in three independent data sets, namely a training set, a validation set, and a test set. Each set contained data from approximately one-third of the participants. Participants were randomly assigned to one of the sets, stratified by both group (i.e., diagnosis) and gender (Table~\ref{tab:demographics_data}). The training and validation set were used to train an optimal deep learning model, whereas the test set was used to evaluate the model performance on \emph{unseen} data (that is, the test set data were not used to train or find the model architecture). Any results on the model performance therefore, were only based on the test set (Section~\ref{sec:Results}).

@@

\subsection{Model \label{subsec:Model}}
\subsubsection{Model Architecture \label{subsubsec:Model_Architecture}}
\begin{figure}[H]
	\centering
	\includegraphics[width=13.5cm]{fig/methods_tcn_arch.pdf}
	\caption{The generic model architecture of the deep learning model to predict initial contacts (ICs) and final contacts (FCs). The inputs are the accelerometer and gyroscope data from a single inertial measurement unit, and are fed to a temporal convolutional network (TCN) (left). The TCN consisted of repeating residual blocks (ResBlock) with exponentially increasing dilation factor (middle). Each ResBlock was built from two sequences of a convolutional layer (Conv), batch normalization layer (BatchNorm), a rectified linear unit activation layer (ReLU), and a dropout layer (DropOut) (right). \label{fig:methods_tcn_arch}}
\end{figure} 
The generic architecture for the deep learning model was based on a temporal convolutional network (TCN) \cite{Bai2018,Remy2020,Filtjens2020}. The TCN consisted of a sequence of residual blocks with exponentially increasing dilation factor \cite{YuKoltun2016,Bai2018}. Each residual block was built from two sequences of a dilated convolutional layer \cite{YuKoltun2016}, a batch normalization layer \cite{Ioffe2015}, a rectified linear unit (ReLU) activation layer, and dropout layer \cite{Srivastava2014} (Figure~\ref{fig:methods_tcn_arch}). The model was built in Python \cite{VanRossum2009} using the high-level TensorFlow API Keras \cite{Chollet2015,Remy2020}.

For the current analysis only convolutions of the "same" type were considered \cite{Remy2020}, i.e., the model was non-causal and zero-padded to account for edge effects, and the likelihood of an IC or FC was based on input data both before and after the current sample, $n$:
\begin{equation}
	\hat{y}_{i}[n] = f(\cdots,\mathbf{X}[n-1],\mathbf{X}[n],\mathbf{X}[n+1],\cdots), \quad i \in \left\{ \mathrm{IC}, \mathrm{FC}\right\}
	\label{eqn:predictions}
\end{equation}
The number of samples that the predictions at time $n$ "sees", was referred to as the \emph{receptive field} \cite{Lea2017} and was a function of the kernel size and the dilation factors \cite{Remy2020}. Dilation factors were always given as a sequence of increasing powers of 2 \cite{VanDenOord2016,YuKoltun2016,Bai2018}.

The outputs of the TCN block were fed to two separate fully-connected (FCN, or dense) layers, that both were followed by a sigmoid activation layer. Outputs were then predicted separately for ICs and FCs. The mean squared error (MSE) was used as loss function, and a gradient descent-based optimization algorithm with adaptive moment (Adam) optimizer was used to iteratively learn the weights \cite{Kingma2014,Schmidt2021}.

\subsubsection{Hyperparameter Optimization}
In order to find the best model architecture, hyperparameter tuning was perfomed using KerasTuner \cite{OMalley2019}. Here, the number of filters, the kernel size, and the maximum dilation factor (Table \ref{tab:hyperparameter_optimization}) were optimized using a random search strategy \cite{Bergstra2012}.

\begin{table}[H] 
	\caption{Model hyperparameters that were optimized for, and the corresponding sets of possible values.\label{tab:hyperparameter_optimization}}
	\newcolumntype{C}{>{\centering\arraybackslash}X}
	\begin{tabularx}{\textwidth}{CC}
		\toprule
		\textbf{Description}	& \textbf{Possible values}\\
		\midrule
		Number of filters	& 8, 16, 32, 64, 128\\
		Kernel size			& 3, 5, 7\\
		Dilations 			& [1, 2], [1, 2, 4], [1, 2, 4, 8]\\
		\bottomrule
	\end{tabularx}
\end{table}

The model architecture that resulted from the hyperparameter optimization was then trained on the combined set of training and validation data. The trained model was used to predict the likelihoods of ICs and FCs events on the hold-out test set.

>>>>

% Authors, for the paper (add full first names)
\Author{Robbin Romijnders$^{1}$*\orcidA{}, Elke Warmerdam$^{2}$, Clint Hansen$^{1}$, Gerhard Schmidt$^{3}$ and Walter Maetzler$^{1}$}

@@

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{%
$^{1}$ \quad Department of Neurology, Kiel University, 24105 Kiel, Germany; r.romijnders@neurologie.uni-kiel.de (R.R.), c.hansen@neurologie.uni-kiel.de (C.H.), w.maetzler@neurologie.uni-kiel.de (W.M.)\\
$^{2}$ \quad Innovative Implant Development (Fracture Healing), Division of Surgery, Saarland University, 66421 Homburg, Germany; elke.warmerdam@uni-saarland.de (E.W.)\\
$^{3}$ \quad Institute of Electrical Engineering and Information Technology, Faculty of Engineering, Kiel University, 24143 Kiel, Germany; gus@tf.uni-kiel.de (G.S.)}

@@

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{Many algorithms use 3D accelerometer and/or gyroscope data from inertial measurement unit (IMU) sensors to detect gait events (i.e., initial and final foot contact). However, these algorithms often require knowledge about sensor orientation and use empirically derived thresholds. As sensor-to-segment alignment cannot always be controlled for in ambulatory assessments, methods are needed that require little knowledge on sensor location and orientation, e.g., a convolutional neural network-based deep learning model. Therefore, 160 participants from healthy and neurologically diseased cohorts walked 5 meter distances at slow, preferred, and fast walking speed, while data were collected from IMUs on the left and right ankle and shank. Gait events were detected and stride parameters were extracted using a deep learning model and an optoelectronic motion capture (OMC) system for reference. Results showed a high detection rate for both initial and final contacts across sensor locations (recall $\ge 92\%$, precision $\ge 97\%$). Time agreement was excellent as witnessed from the median time error (0.005 s) and corresponding inter-quartile range (0.020 s). The extracted stride-specific parameters were in good agreement with parameters derived from the OMC system (maximum mean difference 0.003 s, and corresponding maximum limits of agreement (-0.049 s, 0.051 s) for a $95\%$ confidence level.) Thus, the deep learning approach was considered a valid approach for detecting gait events and extracting stride-specific parameters with little knowledge on exact IMU location and orientation in conditions with and without walking pathologies due to neurological diseases.}

% Keywords
\keyword{gait; gait events; interial measurement unit; deep learning}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
Gait deficits are common in older adults and possibly reflect the presence of an underlying neurodegenerative disease \cite{Snijders2007,Hodgins2008}. For example, conversion to Parkinson's Disease \cite{DelDin2019} or conversion from mild cognitive impairment to Alzheimer's Disease \cite{Koenig2017,Bertoli2018} are linked with changes in spatiotemporal gait parameters. Similarly, temporal gait parameters are different for stroke patients \cite{SchroederVon1995,Mohan2021} and patients with multiple sclerosis \cite{Griskevicius2016,Flachenecker2019} when compared to healthy controls. To objectively quantify gait deficits, stride-specific parameters such as stride time or stride length are often used \cite{Hannink2016}. The beginning and end of a stride are determined from two successive initial contacts (ICs) of the same foot \cite{Perry2010,Whittle2012}. The IC is when the foot contacts the ground and together with the instant at which the foot leaves the ground (final contact, FC), each stride can be divided in a stance and swing phase \cite{Rueterbories2010,Bruening2014}.  The events of IC and FC, also referred to as \emph{gait events}, are commonly determined using force or pressure measuring devices \cite{Bruening2014}, or marker-based optoelectronic motion capture systems (henceforth referred to as the \emph{marker-based} system or method) \cite{Chiari2005,Topley2020}. These systems are relatively expensive, and restricted to usage in expertise laboratories \cite{Iosa2016,Jarchi2018}. As there is increasing evidence that gait measured in the lab does not reflect daily-life gait \cite{Hillel2019,Warmerdam2020,Atrsaei2021}, there is increasingly more interest in measurement systems that allow for continuous gait analysis in ambulatory settings. Therefore, the use of inertial measurement units (IMUs) is especially attractive, as these can be used to measure gait in ecologically valid environments, such as the home environment, thereby painting a more complete picture of health status \cite{DelDin2016,Shah2020} and providing clinical information that is complementary to standardized lab-based assessments \cite{Warmerdam2020,Fasano2020,Atrsaei2021,Corra2021}.

Previous research suggests that gait event detection is more accurate using an IMU worn on a lower limb (e.g., shank or foot) compared to an IMU worn on the low back \cite{BenMansour2015,Storm2016,Panebianco2018}. Now, in order to get from abstract IMU sensor readings to clinically relevant gait parameters (e.g., from accelerations and angular velocities to stride times) \cite{Hannink2016}, different algorithmic approaches have been developed in the last twenty years of clinical gait research. A recent study has evaluated a cross-section of these algorithms for different sensor locations on the lower leg and foot \cite{Niswander2021}. The algorithms were categorized according to which signals were analyzed, for example the angular velocity about the medio-lateral axis, or the accelerations along vertical and antero-posterior axes. This means the sensor readings need to be linked with the anatomical axes, that is, one needs to know which sensor axis aligns with for example the medio-lateral axis. In most approaches, it is simply assumed that due to sensor attachment the sensor axis aligns roughly with the anatomical axis of interest (\cite{Salarian2004,Sabatini2005,Jasiewicz2006,Catalfamo2010,Trojaniello2014,Maqbool2016,Romijnders2021}) or an additional calibration procedure (e.g., \cite{Ferraris1995}) is required (\cite{Greene2010,Niswander2021}). In ambulatory assessments however, study participants often attach the sensor themselves, for example after showering, and therefore the sensor location and alignment cannot be controlled for. Furthermore, it is unlikely that each time the sensor is (re-)attached study participants, especially those with gait deficits, perform a calibration procedure that usually consists of holding a pre-defined pose and performing some known movement sequences \cite{Leineweber2019,Pacher2020}.

Taken together, this drives the need for an approach that is invariant to sensor orientation, and is applicable across a variety of pathological gait patterns. In the field of image analysis similar requirements have been successfully addressed by algorithms that share a common underlying methodology referred to as deep learning \cite{LeCun2015,Hannink2016,TerHaarRomeny2019}. Recent applications of deep learning algorithms have already shown improved performance in detecting gait events from marker-based motion capture when compared to conventional, often heuristics-based, algorithms \cite{Kidzinski2019,Lempereur2020,Filtjens2020}. Another study used a deep learning approach to detect gait events from either three IMUs (worn on the low back, and both ankles) or a single IMU (worn on the low back), and showed that the time error was considerably smaller for the deep learning algorithm than for a commonly applied wavelet-based approach \cite{Gadaleta2019}.

The aim of this study is to extend these works by validating this deep learning approach for detecting gait events in a heterogeneous cohort of healthy and neurologically diseased adults considering a single IMU setup, worn on the lower leg.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Materials and Methods}
\subsection{Data Collection}
Gait analyses were performed in the Universit\"{a}tsklinikum Schleswig-Holstein (UKSH) campus Kiel, Germany. The study \cite{Warmerdam2021} was approved by the ethical committee of the medical faculty at the UKSH (no: D438/18). In total, data from 160 participants were included for the current analysis, including data from young adults (YA; age: 18 - 60 years), older adults (OA; age: $>$60 years), people with Parkinson's Disease (PD; according to the UK Brain Bank criteria \cite{Gibb1988}), people with a recent ($<$4 weeks) symptomatic stroke (stroke), people with multiple sclerosis (MS; according to the McDonalds criteria \cite{Thompson2018}), people with chronic low back pain (cLBP), and people with diagnoses not fitting in any before-mentioned groups or disorders with no explicit diagnosis (other) (Table \ref{tab:demographics_data}). Inclusion criteria were an age of 18 years or older, and the ability to walk independently without a walking aid. Participants were excluded from the study with a Montreal Cognitive Assessment \cite{Nasreddine2005} score $<$15 and other movement disorders that affected mobility, as noticed by the clinical assessor.

@@

\begin{figure}[H]
	\begin{adjustwidth}{-\extralength}{0cm}
		\centering
		\includegraphics[width=13.5cm]{fig/methods_overall_workflow.pdf}
	\end{adjustwidth}
	\caption{Schematic depiction$^{1}$ of the current study. Study participants wore IMUs on the ankle and shanks, and reflective markers were adhered on the heel and toe of usual footwear (illustrated on the left). Marker data were used to obtain reference values for the timings of initial and final contacts (top), where accelerometer and gyroscope data from each tracked point were inputted to a neural network that predicted timings of the same initial and final contacts (bottom). \newline \noindent{\footnotesize{\textsuperscript{1} Picture from: \url{https://www.vecteezy.com/free-vector/man-walking }.}} \label{fig:methods_marker_vs_imu}}
\end{figure}

@@

For the current analysis data from four IMUs (Noraxon USA Inc., myoMOTION, Scottsdale, AZ, USA) were considered, namely those that were attached laterally above the left and right ankle joint and those attached proximally at the left and right shank. IMUs were secured to participants using elastic bands with a special hold for the IMU. Furthermore, reflective markers were attached on top of the usual foot wear at the heel and toe of both feet (Figure~\ref{fig:methods_marker_vs_imu}). Marker data were recorded using a twelve-camera optoelectronic motion capture system (Qualisys AB, G\"{o}teborg, Sweden) at a sampling frequency of 200 Hz. IMU data were recorded at the same sampling frequency, and both systems were synchronized using a TTL signal \cite{Warmerdam2021}.

\subsection{Data Pre-processing \label{subsec:data_preprocessing}}
\subsubsection{Marker Data}
First, data from both marker and IMU systems were cropped so that only data from within the 5 meter distance were considered. Any gaps in the marker data were filled by interpolation making use of inter-correlations between markers \cite{Federolf2013,Gloersen2016}. The data were then low-pass filtered using a 6th order Butterworth filter with a cut-off frequency of 20 Hz \cite{Racz2021}. The filter was applied twice to the input data \cite{Kormylo1974}. After filtering in forward direction, the filtered sequence was reversed and run back through the filter \cite{Salarian2004}. The filtered data were differentiated to get velocity signals, and timings of ICs and FCs were determined from local maxima and minima in the heel and toe vertical velocity signals \cite{Pijnappels2001,OConnor2007}. All identified ICs and FCs were manually checked using Qualisys Track Manager 2018.1 software (Qualisys AB, Göteborg, Sweden), and corrected if necessary \cite{Carcreff2018,Romijnders2021}. The resulting annotated ICs and FCs were considered the \emph{true} events (also \emph{labels} or \emph{targets}), and were used as reference timings to derive stride-specific gait parameters.

\subsubsection{IMU Data}
The idea behind the deep learning approach was that a \emph{model} was trained to predict the likelihood of an IC and FC, given accelerometer and gyroscope data from a single IMU. The data from a single sensor channel, e.g., the acceleration in forward direction, was denoted by $\mathbf{x}_{d} = \begin{bmatrix}
	x_{d}[1] & x_{d}[2] & \cdots & x_{d}[N]
\end{bmatrix}^{\mathrm{T}}$, with $d$ referring to the $d$th sensor channel (i.e., $d = 1, \cdots, D$) and $n$ referring to the $n$th sample or time step (i.e., $n = 1, \cdots, N$). Similarly, the data from all $D$ sensor channels at a given time instant $n$, was denoted by $\mathbf{x}[n] = \begin{bmatrix}
x_{1}[n] & x_{2}[n] & \cdots & x_{D}[n]
\end{bmatrix}^{\mathrm{T}}$. Data from all $D$ channels, and for all $N$ time steps, was then denoted by:
\begin{equation}
	\mathbf{X} = \begin{bmatrix}
		\rule[0ex]{0.5pt}{2.5ex} & \rule[0ex]{0.5pt}{2.5ex} &  & \rule[0ex]{0.5pt}{2.5ex}\\
		\mathbf{x}_{1} & \mathbf{x}_{2} & \cdots & \mathbf{x}_{D}\\
		\rule[0ex]{0.5pt}{2.5ex} & \rule[0ex]{0.5pt}{2.5ex} &  & \rule[0ex]{0.5pt}{2.5ex}
	\end{bmatrix} = \begin{bmatrix}
		x_{1}[1] & x_{2}[1] & & x_{D}[1]\\
		x_{1}[2] & x_{2}[2] & & x_{D}[2]\\
		\vdots & \vdots & \cdots & \vdots\\
		x_{1}[N] & x_{2}[N] & & x_{D}[N]\end{bmatrix}, \quad \mathbf{X} \in \mathbb{R}^{N \times D}
	\label{eqn:train_data_single_instance}
\end{equation}

Likewise, the \emph{labels} were denoted by:
\begin{equation}
	\mathbf{y}_{\mathrm{IC}} = \begin{bmatrix}
		y_{\mathrm{IC}}[1]\\
		y_{\mathrm{IC}}[2]\\
		\vdots\\
		y_{\mathrm{IC}}[N]
	\end{bmatrix}, \quad \mathbf{y}_{\mathrm{FC}} = \begin{bmatrix}
		y_{\mathrm{FC}}[1]\\
		y_{\mathrm{FC}}[2]\\
		\vdots\\
		y_{\mathrm{FC}}[N]
	\end{bmatrix}, \quad y_{\mathrm{IC}/\mathrm{FC}}[n] \in [0, 1]
	\label{eqn:train_labels_single_instace}
\end{equation}

The model was iteratively trained to learn a mapping $h_{\mathbf{\Theta}}(\mathbf{X}): \mathbf{X} \rightarrow \mathbf{y}$, where $h_{\mathbf{\Theta}}$ was also referred to as the \emph{hypothesis} parameterized by the weights, collectively denoted by $\mathbf{\Theta}$.

All participant data data were split in three independent data sets, namely a training set, a validation set, and a test set. Each set contained data from approximately one-third of the participants. Participants were randomly assigned to one of the sets, stratified by both group (i.e., diagnosis) and gender (Table~\ref{tab:demographics_data}). The training and validation set were used to train an optimal deep learning model. Test set data were not used for model training or hyperparameter tuning. The results on the model's performance were only based on the test set, and therefore reflect how good the model generalizes to new, \emph{unseen} data.

@@

\subsection{Model}
\subsubsection{Model Architecture}
The generic architecture for the deep learning model was based on a temporal convolutional network (TCN) \cite{Bai2018,Remy2020,Filtjens2020}. The TCN consisted of a sequence of residual blocks with exponentially increasing dilation factor \cite{YuKoltun2016,Bai2018}. Each residual block was built from two sequences of a dilated convolutional layer \cite{YuKoltun2016}, a batch normalization layer \cite{Ioffe2015}, a rectified linear unit (ReLU) activation layer, and a dropout layer \cite{Srivastava2014} (Figure). The model was built in Python \cite{VanRossum2009} using the high-level TensorFlow API Keras \cite{Remy2020,Chollet2015}.

For the current analysis only convolutions of the "same" type were considered \cite{Remy2020}, i.e., the model was non-causal and zero-padded to account for edge effects, and the likelihood of an IC or FC was based on input data both before and after the current sample, $n$:
\begin{equation}
	\hat{y}_{i}[n] = f\left(\cdots, \mathbf{x}[n-1], \mathbf{x}[n], \mathbf{x}[n+1], \cdots \right), \quad i \in \left\{ \mathrm{IC}, \mathrm{FC}\right\}
\end{equation}
The number of samples that the predictions at time $n$ "sees", was referred to as the \emph{receptive field} \cite{Lea2017} and was a function of the kernel size and the dilation factors \cite{Remy2020}. Dilation factors were always given as a sequence of increasing power of 2 \cite{YuKoltun2016,Bai2018,VanDenOord2016}.

The outputs of the TCN block were fed to two separate fully-connected (FCN, or dense) layers, that both were followed by a sigmoid activation layer. Outputs were then predicted separately for ICs and FCs \cite{Gadaleta2019,Filtjens2020}. The mean squared error (MSE) was used as a loss function, and a gradient descent-based optimization algorithm with adaptive moment (Adam) optimizer was used to iteratively learn the weights \cite{Kingma2014,Schmidt2021}.

\subsubsection{Hyperparameter Optimization}
In order to find the best model architecture, hyperparameter tuning was perfomed using KerasTuner \cite{OMalley2019}. Here, the number of filters, the kernel size, and the maximum dilation factor (Table \ref{tab:hyperparameter_optimization}) were optimized for using a random search strategy \cite{Bergstra2012}.

\begin{table}[H] 
	\caption{Model hyperparameters that were optimized for, and the corresponding sets of possible values.\label{tab:hyperparameter_optimization}}
	\newcolumntype{C}{>{\centering\arraybackslash}X}
	\begin{tabularx}{\textwidth}{CC}
		\toprule
		\textbf{Description}	& \textbf{Possible values}\\
		\midrule
		Number of filters	& 8, {\bf 16}, 32, 64, 128\\
		Kernel size			& 3, {\bf 5}, 7\\
		Dilations 			& [1, 2], {\bf [1, 2, 4]}, [1, 2, 4, 8]\\
		\bottomrule
	\end{tabularx}
	\noindent{\footnotesize{In bold typesetting the hyperparameter values that were selected for the trained model to make predictions on the test set.}}
\end{table}

The model architecture that resulted from the hyperparameter optimization was then trained on the combined set of training and validation data. The trained model was used to predict the likelihoods of ICs and FCs on the test set data.
